Testing and Isolation Efficacy: Insights from a Simple Epidemic Model

### 2021-05-06 SIR_manuscript.tex
analysis, R0 wrt omega

{\bf The second goal} is to explore how changes in \percap testing intensity $\rho$ affects $\Rnum$. Mathematically we would like to verify the sign of $\pder[\Rnum]{\rho}$, which specifically depends on $\pder[\Delta]{\rho}$. We use the derived expressions for $S\_u^*$ and $S\_n^*$, given by Eqs.\,\eqref{dfe}, in $\Delta$ \eqref{eq:del4_theta}. Also, we define $\phi = \hat F_S = \frac{\rho \omega}{\omega-\rho}$, to reparametrize $\rho$. This is mainly to avoid singularity in $\hat F_I$ \eqref{eq:fi}, when testing intensity $\rho$ is very close to the rate of test return $\omega$. Thus, $\rho$ is reparametrized as 
\begin{equation}
\label{eq:phi}
\rho=\frac{\omega \phi}{\omega+\phi}.
\end{equation}
This one-to-one monotonic reparametrization enables us to simplify the mathematical expressions and explore the simpler $\pder[\Delta]{\phi}$ instead of the complicated $\pder[\Delta]{\rho}$.
% with an equivalent but simplified $\Delta$ in Eq.\,\eqref{eq:abc} as
% \begin{equation}
% \label{eq:del3}
% \Delta= \frac{K_1\,\phi^2+K_2\,\phi}{K_3},
% \end{equation}
% where
% \begin{align}
% \label{eq:K}
% K_1=& \frac{w_I}{w_S}\Big(((1+p_I)\omega+\gamma)\gamma\theta\_w+\omega^2 p_I\theta\_c \Big), \\
% K_2=& (\omega+\gamma)\Big( (\frac{w_I}{w_S}\omega+\gamma)\gamma\theta\_w + \frac{w_I}{w_S} \omega^2 p_I\theta\_c \Big),\\
% K_3=& (\omega+\gamma) \Big((\omega p_I+\gamma)\frac{w_I}{w_S}\phi\,+\,(\omega+\gamma)\gamma \Big)(\omega+\phi).
% \end{align}
The derivative is
\begin{align}
\label{eq:dd3dr}
\partial\Delta/\partial\phi=& \frac{1}{d_3} (a_3\,\phi^2+b_3\,\phi+c_3),
\end{align}
where

\begin{multline}
\label{eq:a3}
a_3 = \frac{w_I}{w_S}
\Bigg(\Big(
( \theta_{{w}}{{\it p\_I}}^{2}-\theta_{{c}}{{\it p\_I}}^
{2}\theta_{{w}} ) {\omega}^{3} 
+ ( 2\,\theta_{{w}}{{\it p\_I
}}^{2}-\theta_{{c}}{{\it p\_I}}^{2}-{\it p\_I}\,{\theta_{{w}}}^{2}-
\theta_{{c}}{\it p\_I}\,\theta_{{w}}-{{\it p\_I}}^{2}{\theta_{{w}}}^{2
}+2\,\theta_{{w}}{\it p\_I} ) \gamma\,{\omega}^{2} \\
 + ( -2\,{ \it p\_I}\,{\theta_{{w}}}^{2}-\theta_{{c}}{\it p\_I}+3\,\theta_{{w}}{
\it p\_I}-{\theta_{{w}}}^{2}+\theta_{{w}} ) {\gamma}^{2}\omega 
 + ( \theta_{{w}}-{\theta_{{w}}}^{2} ) {\gamma}^{3} \Big) \frac{w_I}{w_S}
 + ( -\theta_{{c}}{\it p\_I}\,\theta_{{w}}+\theta_{{c}}{\it p\_I}
 ) \gamma\,{\omega}^{2}+ \\
 ( \theta_{{w}}+\theta_{{c}}{\it
p\_I}-{\theta_{{w}}}^{2}-\theta_{{c}}{\it p\_I}\,\theta_{{w}} )
{\gamma}^{2}\omega+ ( \theta_{{w}}-{\theta_{{w}}}^{2} ) {
\gamma}^{3}
\Bigg),
\end{multline}
\begin{align}
\label{eq:abcd2}
b_3&= 2\frac{w_I}{w_S}(\omega+\gamma)\gamma\Big( 
(\omega+\gamma+\omega\,p_I)(2-\theta\_w)\,\gamma\,\theta\_w+ (1-\theta\_w)\,\omega^2 p_I\theta\_c+\omega^2 p_I\theta\_w 
\Big) ,\\
c_3&= (\omega+\gamma)^2\,\gamma\Big( 
(2-\theta\_w)\,\gamma^2\,\theta\_w+
(1+\frac{w_I}{w_S})\,\omega\,\gamma\,\theta\_w+
\frac{w_I}{w_S} \omega^2 p_I\theta\_c \Big),\\
d_3&= \frac{(\omega+\gamma)}{\omega} \Big((\omega p_I+\gamma)\frac{w_I}{w_S}\phi\,+\,(\omega+\gamma)\gamma \Big)^2(\omega+\phi)^2. 
\end{align}
Note that $\phi\geq 0$, also $b_3$, $c_3$ and $d_3$ are all positive. However $a_3$ can be positive or negative.
If $a_3\geq 0$, $\partial\Delta/\partial\phi \geq 0$ for all feasible range of parameters, thus $\pro\Rnum \leq 0$. 
If $a_3 < 0$, then the quadratic expression in the numerator of \eqref{eq:dd3dr} has a positive root, $\phi^*$, such that for $\phi>\phi^*$, $\partial\Delta/\partial\phi < 0$. An example of this counter-vailing effect of $\rho$ on $\Rnum$ occures when $\theta\_w=0$ and $\theta\_c=1$. This is illustraited in the top-right panel of the Fig.\,\ref{pan2} panel (b), where the strength of isolation for awaiting people is the least, but the most for the confirmed cases. In this case $$a_3=\frac{w_I}{w_S}\,\omega\,\gamma\,p_I((\omega+\gamma)-\frac{w_I}{w_S}(\omega\,p_I+\gamma)).$$
Specifically, in the case of targeted testing which is identified with $\frac{w_I}{w_S}> 1$, and using a perfect sensative test, thus $p_I=1$, there exists a range for $\rho$ over which $\pder[\Rnum]{\rho}\leq\,0$.  
Note that $\rho$ and $\omega$ have a similar mechanism in delaying people to get into $I\_c$, thus we would expect to see the counter-vailing effect, which is not obvious, of these two parameters on $\Rnum$. 
% that people in the DFE waiting for negative test, $S_n^*$.
% More testing may have a counter effect on $\Rnum$, as if we have more testing we have more people in $S_n^*$ which will take them longer to get to $I\_c$.
\ali{I stoped here!}

% %%%%%%%
\subsection{rate of returning tests} \label{app:omega}
\ali{needs editig}
The linearization of $\Rnum$ around $\rho=0$ is
\begin{equation}\label{linearization}
\Rnum \approx \beta/\gamma + \frac{\beta \rho}{\omega (\omega+\gamma) \gamma^2 w_S} \Big(\gamma(-\theta\_w)(\gamma w_S+\omega w_I) + (-\theta\_c)p_I w_I \omega^2 \Big). 
\end{equation}
So when $\rho \approx 0$ we have $$\partial{\Rnum}/\partial{\omega} \approx  \frac{-\beta \rho}{\gamma w_S\omega^2 (\gamma+\omega)^2}  (a \omega^2 + b \omega + c),$$
where $a=(-\theta\_w)w_I-(-\theta\_c)p_I w_I$, $b=2(-\theta\_w)\gamma w_S$ and $c=(-\theta\_w)\gamma^2 w_S$.

>>>>

{\bf The second goal} is to explore how changes in \percap testing intensity $\rho$ affects $\Rnum$. Mathematically we would like to verify the sign of $\pder[\Rnum]{\rho}$, which specifically depends on $\pder[\Delta]{\rho}$. We use the derived expressions for $S\_u^*$ and $S\_n^*$, given by Eqs.\,\eqref{dfe}, in $\Delta$ \eqref{eq:del4_theta}. Also, we define $\phi = \hat F_S = \frac{\rho \omega}{\omega-\rho}$, to reparametrize $\rho$. This is mainly to avoid singularity in $\hat F_I$ \eqref{eq:fi}, when testing intensity $\rho$ is very close to the rate of test return $\omega$. Thus, $\rho$ is reparametrized as 
\begin{equation}
\label{eq:phi}
\rho=\frac{\omega \phi}{\omega+\phi}.
\end{equation}
This one-to-one monotonic reparametrization enables us to simplify the mathematical expressions and explore the simpler $\pder[\Delta]{\phi}$ instead of the complicated $\pder[\Delta]{\rho}$.
% with an equivalent but simplified $\Delta$ in Eq.\,\eqref{eq:abc} as
% \begin{equation}
% \label{eq:del3}
% \Delta= \frac{K_1\,\phi^2+K_2\,\phi}{K_3},
% \end{equation}
% where
% \begin{align}
% \label{eq:K}
% K_1=& \frac{w_I}{w_S}\Big(((1+p_I)\omega+\gamma)\gamma\theta\_w+\omega^2 p_I\theta\_c \Big), \\
% K_2=& (\omega+\gamma)\Big( (\frac{w_I}{w_S}\omega+\gamma)\gamma\theta\_w + \frac{w_I}{w_S} \omega^2 p_I\theta\_c \Big),\\
% K_3=& (\omega+\gamma) \Big((\omega p_I+\gamma)\frac{w_I}{w_S}\phi\,+\,(\omega+\gamma)\gamma \Big)(\omega+\phi).
% \end{align}
The derivative is
\begin{align}
\label{eq:dd3dr}
\partial\Delta/\partial\phi=& \frac{1}{d_3} (a_3\,\phi^2+b_3\,\phi+c_3),
\end{align}
where
\begin{multline}
\label{eq:a3}
a_3 = \frac{w_I}{w_S}
\Bigg(\Big(
( \theta_{{w}}{{\it p\_I}}^{2}-\theta_{{c}}{{\it p\_I}}^
{2}\theta_{{w}} ) {\omega}^{3} 
+ ( 2\,\theta_{{w}}{{\it p\_I
}}^{2}-\theta_{{c}}{{\it p\_I}}^{2}-{\it p\_I}\,{\theta_{{w}}}^{2}-
\theta_{{c}}{\it p\_I}\,\theta_{{w}}-{{\it p\_I}}^{2}{\theta_{{w}}}^{2
}+2\,\theta_{{w}}{\it p\_I} ) \gamma\,{\omega}^{2} \\
 + ( -2\,{ \it p\_I}\,{\theta_{{w}}}^{2}-\theta_{{c}}{\it p\_I}+3\,\theta_{{w}}{
\it p\_I}-{\theta_{{w}}}^{2}+\theta_{{w}} ) {\gamma}^{2}\omega 
 + ( \theta_{{w}}-{\theta_{{w}}}^{2} ) {\gamma}^{3} \Big) \frac{w_I}{w_S}
 + ( -\theta_{{c}}{\it p\_I}\,\theta_{{w}}+\theta_{{c}}{\it p\_I}
 ) \gamma\,{\omega}^{2}+ \\
 ( \theta_{{w}}+\theta_{{c}}{\it
p\_I}-{\theta_{{w}}}^{2}-\theta_{{c}}{\it p\_I}\,\theta_{{w}} )
{\gamma}^{2}\omega+ ( \theta_{{w}}-{\theta_{{w}}}^{2} ) {
\gamma}^{3}
\Bigg),
\end{multline}
\begin{align}
\label{eq:abcd2}
b_3&= 2\frac{w_I}{w_S}(\omega+\gamma)\gamma\Big( 
(\omega+\gamma+\omega\,p_I)(2-\theta\_w)\,\gamma\,\theta\_w+ (1-\theta\_w)\,\omega^2 p_I\theta\_c+\omega^2 p_I\theta\_w 
\Big) ,\\
c_3&= (\omega+\gamma)^2\,\gamma\Big( 
(2-\theta\_w)\,\gamma^2\,\theta\_w+
(1+\frac{w_I}{w_S})\,\omega\,\gamma\,\theta\_w+
\frac{w_I}{w_S} \omega^2 p_I\theta\_c \Big),\\
d_3&= \frac{(\omega+\gamma)}{\omega} \Big((\omega p_I+\gamma)\frac{w_I}{w_S}\phi\,+\,(\omega+\gamma)\gamma \Big)^2(\omega+\phi)^2. 
\end{align}
Note that $\phi\geq 0$, also $b_3$, $c_3$ and $d_3$ are all positive. However $a_3$ can be positive or negative.
If $a_3\geq 0$, $\partial\Delta/\partial\phi \geq 0$ for all feasible range of parameters, thus $\pro\Rnum \leq 0$. 
If $a_3 < 0$, then the quadratic expression in the numerator of \eqref{eq:dd3dr} has a positive root, $\phi^*$, such that for $\phi>\phi^*$, $\partial\Delta/\partial\phi < 0$. An example of this counter-vailing effect of $\rho$ on $\Rnum$ occures when $\theta\_w=0$ and $\theta\_c=1$. This is illustraited in the top-right panel of the Fig.\,\ref{pan2} panel (b), where the strength of isolation for awaiting people is the least, but the most for the confirmed cases. In this case $$a_3=\frac{w_I}{w_S}\,\omega\,\gamma\,p_I((\omega+\gamma)-\frac{w_I}{w_S}(\omega\,p_I+\gamma)).$$
Specifically, in the case of targeted testing which is identified with $\frac{w_I}{w_S}> 1$, and using a perfect sensative test, thus $p_I=1$, there exists a range for $\rho$ over which $\pder[\Rnum]{\rho}\leq\,0$.  
Note that $\rho$ and $\omega$ have a similar mechanism in delaying people to get into $I\_c$, thus we would expect to see the non-trivial counter-vailing effect of these two parameters on $\Rnum$.

% %%%%%%%
\subsection{rate of returning tests} \label{app:omega}
{\bf The third goal} is to explore how changes in the rate of test return $\omega$ affects $\Rnum$. Mathematically we would like to verify the sign of $\pder[\Rnum]{\omega}$, which specifically depends on $\pder[\Delta]{\omega}$. We use
the linearization of $\Rnum$ around $\rho=0$ to show that there a non-monotonic relationship between $\Rnum$  and $\omega$. The taylor expansion of $\Delta$ at $\rho=0$ is
\begin{equation}
\label{eq:lin}
\Delta = \frac{\rho}{\omega\,\gamma(\omega+\gamma)} \Big(
\frac{w_I}{w_S}\,\omega^2\,p_I\,\theta\_c+(\frac{w_I}{w_S}+1)\,\gamma\omega\theta\_w\,+\gamma^2\,\theta\_w(2-\theta\_w)
\Big). 
\end{equation}
This results in
\begin{equation}
\label{eq:dlindo}
\pder[\Delta]{\omega} = \frac{\rho}{\omega^2\,(\omega+\gamma)^2} \Big(
(p_I\,\frac{w_I}{w_S}\theta\_c\,-(1+\frac{w_I}{w_S})\theta\_w)\omega^2+2\theta_w\,\gamma(\theta_w-2)\,\omega+\theta_w\,\gamma^2\,(\theta_w-2)
\Big),
\end{equation}
around $\rho=0$.
\ali{I stoped here!}
